System for activity tracking of patients with chronic kidney disease

### 2015-07-07 MScCS.tex
jens ändrade

\usepackage{hyperref}

\author{
	Jens Gustafsson \\
	{\normalsize \href{mailto:ada10jgu@student.lu.se}{\texttt{ada10jgu@student.lu.se}}}
	\and
	Alfred Åkesson \\
    {\normalsize \href{mailto:ada10aak@student.lu.se}{\texttt{ada10aak@student.lu.se}}}
}
\title{System for activity tracking of patients with kidney failure}

%\title{RSAT}
%\subtitle{A system for activity tracking of patients with kidney failure.}
%\company{The Corporation AB LTD Inc}
\supervisors{Boris Magnusson, \href{mailto:Boris.Magnusson@cs.lth.se}{\texttt{Boris.Magnusson@cs.lth.se}}}{Fredrik Nilsask, \href{mailto:fredrik.nilsask@cs.lth.se}{\texttt{fredrik.nilsask@cs.lth.se}}}
\examiner{Görel Hedin, \href{mailto:gorel.hedin@cs.lth.se}{\texttt{gorel.hedin@cs.lth.se}}}

@@

\acknowledgements{
TODO!
%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.

@@

\theabstract{
Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. Regular exercise and physical activity are therefore crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help patients with kidney failure to lose weight. A question they asked themselves was whether it was possible to use activity tracking devices in the project. The purpose of this thesis is to design and evaluate a system that can be used to track the physical activities of the patients. The system will be built using a Sony SmartBand, a wristband collecting and analyzing data about a user's daily activities.
}

\keywords{kidney failure, activity tracking, lose weight, Sony Lifelog}

@@

Many people suffering from kidney failure are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. It is very hard for this group of people to lose weight, as further described in section \ref{sec:kiney}. Regular exercise and physical activity are therefore crucial for these patients.

This thesis presents a system with the aim of helping over-weight patients with kidney failure to lose weight. The system is called the Region Skåne Activity Tracking, RSAT. 
itACiH is a research project that investigates the possibilities of advanced health care at home. As a part of the itACiH project a planning-system was developed that is currently being used at the hospitals in Skåne. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. RSAT is thought to be integrated with the planning system. 
%% Förslag på ändring
% RSAT was developed to be integrated in a planning system already in use at the hospital. The planning system is develop in a research project called itACiH. itACiH investigates the possibilities of advanced health care at home. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. 
A project group at the Skåne University Hospital (called the RSAT-group in this thesis) has recently been established to help patients in need of a kidney transplant to lose weight. One of their wishes was to use activity tracking devices in the project.

Sony has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their cloud and accessible through an API called the Lifelog API. Sony wants to investigate what different use cases that exists for their wearable devices and has chosen to sponsor this project with smart wristbands for the patients to wear.

\newpage

\section{Purpose and question formulation}

The purpose of this thesis is to build a system that can be used in the health care to help patients with kidney failure to lose weight. The Sony Lifelog API will be used as the foundation for the system. As a step in building the system, an investigation about how well suited the API is to be used will be performed.

@@

\begin{quote}
\centering
\emph{How well suited is the Sony Lifelog platform to be used in the heath care from a technical standpoint?}
\end{quote}

\begin{quote}
\centering
\emph{How well suited is RSAT, the system to be developed, to be used in the health care?}
\end{quote}

\begin{comment}
The following questions will be used to fulfill the purpose:

\begin{quote}
\centering
\emph{In what ways can the Sony Lifelog technology be used in the health care to help patients with renal failure?}
\end{quote}

This is a very broad question, and it has therefore been divided into three sub questions:

\begin{enumerate}
    \item \emph{What other platforms exists on the market that could have been used instead of the Sony Lifelog?}
    \item \emph{What are the pros and cons of using the Sony Lifelog API?}
    %\item \emph{}
    %\item \emph{What data in the Lifelog API is useful for the health care?}
    %\item \emph{What functionality does not exist in the Sony Lifelog technique which would be useful for the health care?}
    \item \emph{What problems, if any, do arise when third party API:s are integrated in the health care, from a technical standpoint?}
\end{enumerate}

The system will be made up from requirements that have been created in collaboration with a physiotherapist. The physiotherapist is working with patients suffering from chronic kidney disease on a daily basis and is the one who will be using the system. These requirements will be presented in \ref{sec:req}.

\end{comment}
\section{Method}
Before the development of RSAT could start, an interview with the RSAT-group was held. Their wishes was used as input when the requirements was created. Thereafter a proof-of-concept system was developed to verify that it was possible to use the Lifelog API.

Iterative prototyping was used to develop RSAT. For each iteration, the current requirements was identified and implemented. Each iteration resulted in a demo that the RSAT-group evaluated. Feedback from the RSAT-group was used as input to the next iteration. Sometimes the feedback resulted in new requirements on the system and if those was feasible, they was added to a development iteration.

@@

%%%%%%%%%%%%%%%%%%

\section{Disposition}
The thesis is organized into three different parts; an overview part, a development part and an evaluation part.

The overview part consists of chapters 2, 3 and 4. It gives a medical background, describing why RSAT is needed and a technical background that describes the Lifelog platform. It also contains an introduction to PalCom, the framework that RSAT is built on. The last chapter in this part describes related work. Related work includes information about different wearable technologies, platforms for aggregation of activity data and other systems that are similar to RSAT.

The development part consists of chapters 5 and 6. This part contains information about the requirements on the system, what decisions that were made, why the decisions were made and how RSAT was implemented.

The evaluation part consists of chapters 7 and 8. RSAT and Lifelog are evaluated and compared to the other systems described in the first part of the thesis.

% BRA KÄLLA to be used: http://www.kidney.ca/facing-the-facts

@@

Kidneys have the responsibility of filtering waste products from the blood in the body\cite{NKDEPBASICS}. When the kidney function is too low to filter the blood, dialysis and in the end a kidney transplant, is needed.

In this chapter an introduction to chronic kidney disease is given, beginning with a short introduction to chronic conditions in general. Chapter 2.2 discusses chronic kidney disease especially.

@@

The costs in the health care are ever increasing. A huge part of the health care and the economic costs associated with it addresses patients with chronic conditions. More than 60\% of all deaths in the world is due to this types of complications. In Sweden 80\% of all health care costs go to treating these conditions\cite{SmartCare}. A study shows that three types of lifestyle behaviors  leads to four chronic diseases. These diseases accounts for about 50\% of all the deaths worldwide, see figure \ref{fig:chronic-fig}\cite{callaway2015quantified}.

@@

Physical inactivity, as seen in figure \ref{fig:chronic-fig} , is one factor that could lead to a chronic condition.

\section{Chronic kidney disease}
\label{sec:kiney}
\begin{comment}
NKDEPBASICS: http://nkdep.nih.gov/learn/kidney-disease-basics.shtml
KIDNEYFOUNDATION: https://www.kidney.org/kidneydisease/aboutckd
NKDEPPROG: http://nkdep.nih.gov/identify-manage/manage-patients/slow-progression.shtml
DIALYSIS: http://www.vardhandboken.se/Texter/Dialys-peritonealdialys/Oversikt/
\end{comment}

Chronic kidney disease (CKD) is a gradual loss of kidney function over time\cite{KIDNEYFOUNDATION}. There are no specific symptoms of the disease but illness, tiredness and a reduced appetite are all common symptoms. As the kidney function is worsening, waste products build to high levels in the blood and may cause complications such as high blood pressure,  muscle and bone mass loss. The risk of getting other diseases such as cardiovascular disease, pericarditis and anemia is also growing.  When the waste products in a patient’s body becomes so high that the patient starts to get sick from it, the patient requires dialysis\cite{KIDNEYFOUNDATION}.%todo: kolla över mening

People with severe CKD are in need of a kidney transplant. Until a transplant can be performed, dialysis is required. Without dialysis, the patients would have a poor life expectancy\cite{KIDNEYFOUNDATION}. Dialysis is a cumbersome treatment process for a patient as it takes up a lot of time and adds a lot of constraints to the patient’s life. Dialysis is also expensive for the society. A rough estimate of the annual cost of treating one patient with dialysis in Sweden is 250 000 – 500 000 SEK. \cite{sr}

Peritoneal dialysis and hemodialysis are dialysis that can be performed in the home of a CKD-patient. The principle behind peritoneal dialysis is that a sterile solution containing glucose is run into the patient’s abdomen\cite{DIALYSIS}. The peritoneum inside the patient acts as a membrane across which waste products from the blood is filtered. A downside with the peritoneal dialysis is that the calorie intake increases, since the sterile solution contains glucose. Hemodialysis is a kind of blood dialysis, where blood is filtered outside the body.

Despite that CKD is considered to be an irreversible condition, the progress of the disease can be slowed down. The National Kidney Disease Education Program (NKDEP) lists three strategies that can be used to slow down the progress\cite{NKDEPPROG}. One of those strategies is lifestyle interventions and addresses different health-promoting behaviour. One of the health-promoting behaviours they discuss is physical activity. NKDEP states that people suffering from CKD have the same recommendations for physical activity as the general population.

Eventually, a patient with CKD must undergo a kidney transplant. The patient must be in good physical shape, as the risk of infections and other diseases otherwise are too high. It is also essential that the patient does not have too much belly fat. Not only does the fat induce a greater risk for the patient's life, it also makes it harder for surgeons to practically perform the transplant. %Vad kostar det?!
%http://www.kidney.ca/document.doc?id=4083
%http://health.costhelper.com/kidney-cancer.html

\chapter[Technical Background]{Technical Background}

This chapter describes the techniques used to develop RSAT. First the Sony Lifelog and techniques used by Lifelog are described. The second part describes PalCom which is the platform used by the itACiH and RSAT.

\section{Sony Lifelog}

The Sony Lifelog application is an Android application which is built around the idea of logging a users lifestyle and daily activities\cite{LifeLogDescr}. The application collects various data about its users through different sensors in a smartphone and SmartWear-devices (Sony SmartWatch or a Sony Smartband) and displays it to the user. Example of data it logs \emph{?!?!are/is?!?}  information about what applications in the smartphone that have been used, what locations that have been visited and information about different physical activities performed during the day.

\subsection{Sony SmartBand}

The Sony SmartBand SWR10 is the chosen wristband for this project. The Sony SWR10 has a movement sensor that recognizes different movements a user makes\cite{SWR10}. This data is analyzed by Sony and categorized into different activates, which are uploaded to a central repository that stores the data. The SmartBand must be connected to an Android device with Android version 4.4 or higher to be able to collect data.

\subsection{Sony Lifelog API} 
All the data collected in the Sony Lifelog application are automatically uploaded to a central repository that stores the data. This data is globally accessible through a REST-based web-service API\cite{LifeLogDescr}.

@@

Third parties can access the API to retrieve collected data about a user. Every request made to the API must be authorized and https must be the chosen communication protocol. The lifelog Web API uses OAuth 2.0 as its authorization mechanism. To connect to the API, a user must login and give explicit permission for the application or the service to access to the data.

@@

The Sony Lifelog API is a REST API. REST is an architectural style for designing API:s and are heavily used in modern web services. The concept is to represent resources as the analogy for the data endpoints and to have a small set of actions interacting with these resources. The REST architecture is based on the following principles\cite{pautasso2014restful}:

@@

@@

OAuth 2.0 is an authorization protocol developed by the Internet Engineering Task Force (IETF). It describes an open standard for authorization and enables third party applications to obtain limited access to a web-service on the behalf of a resource owner\cite{oauth2spec}. To describe the protocol, several key actors need to be described, these actors are the client, the resource owner, the authorization server and the resource server. IETF gives the following definitions to the roles:

@@

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed\cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code, which is appended in the URL. The authentication code is short lived and single-use.

@@

\section{Evaluation of RSAT}

>>>>

\usepackage{hyperref}
\usepackage{float}
\restylefloat{table}

\author{
	Jens Gustafsson \\
	{\normalsize \href{mailto:ada10jgu@student.lu.se}{\texttt{ada10jgu@student.lu.se}}}
	\and
	Alfred Åkesson \\
    {\normalsize \href{mailto:ada10aak@student.lu.se}{\texttt{ada10aak@student.lu.se}}}
}
\title{System for activity tracking of patients with chronic kidney disease}

%\title{RSAT}
%\subtitle{A system for activity tracking of patients with kidney failure.}

\supervisor{Boris Magnusson, \href{mailto:Boris.Magnusson@cs.lth.se}{\texttt{Boris.Magnusson@cs.lth.se}}}
\examiner{Görel Hedin, \href{mailto:gorel.hedin@cs.lth.se}{\texttt{gorel.hedin@cs.lth.se}}}

@@

\acknowledgements{
%  Läs och ändra Jens
We want to thank all the people in the project group especially Fredrik the project manager and Mia the physiotherapist. We also want to thank Sony for sponsoring with wristbands. We also wan the acknowledge our supervisor Boris and examiner Görel. Last we want to do a shout out to our siblings in Kids of Palcom.

%TODO!
%If you want to thank people, do it here, on a separate right-hand page. Both the U.S. \textit{acknowledgments} and the British \textit{acknowledgements} spellings are acceptable.

@@

\theabstract{
Many people suffering from chronic kidney disease are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. Regular exercise and physical activity are therefore crucial for this group of people. A project group at the Skåne University Hospital has recently been established to help patients with chronic kidney disease to lose weight. A question they asked themselves was whether it was possible to use activity tracking devices in the project. The purpose of this thesis is to design and evaluate a system that can be used to track the physical activities of the patients. The system will be built using a Sony SmartBand, a wristband collecting and analyzing data about a user's daily activities.
}

\keywords{chronic kidney disease, activity tracking, lose weight, Sony Lifelog}

@@

Many people suffering from chronic kidney disease are in need of a kidney transplant. A problem in the health care is that patients cannot undergo surgery if they have too much belly fat. It is very hard for this group of people to lose weight, as further described in section \ref{sec:kiney}. Regular exercise and physical activity are therefore crucial for these patients.

% TODO VAD ÄR WELFARE OFFCIER?
A project group at the Skåne University Hospital has recently been established to help patients in need of a kidney transplant to lose weight. The group consists of a doctor, a psychotherapist, a dietitian and a welfare officer. One of their wishes was to use activity tracking devices in the project. For this reason a project group established to investigate the possibility's. This group is called the RSAT-group.

This thesis presents a system with the aim of helping over-weight patients with chronic kidney disease to lose weight. The system is called the Region Skåne Activity Tracking, RSAT.

itACiH is a research project that investigates the possibilities of advanced health care at home. As a part of the itACiH project a planning-system was developed that is currently being used at the hospitals in Skåne. The planning system is built as a web application and can fetch data from sensors that are installed in a patient's home. RSAT is intended to be integrated with the planning system.

RSAT was developed in collaboration with the RSAT-group. The focus during the development has therefore been centered around a specific use case. Despite this, RSAT could be used to help other groups of people losing weight or to get in a better physical condition. RSAT also has a very modular architecture, which means that it easily can be adjusted to fit the needs that might evolve in other use cases.

Sony Mobile has developed wearable devices that can track a user’s daily activities. The data collected by these devices are saved online in their cloud and accessible through an API called the Lifelog API. Sony wants to investigate what different use cases that exists for their wearable devices and has chosen to sponsor this project with smart wristbands for the patients to wear.

\section{The RSAT-group}
RSAT has been developed in collaboration with a project group established at the Region Skåne University hospital. In this thesis this group is referred to as the RSAT-group. Sony Mobile also was a member of this group and it was therefore chosen that RSAT should be built around their Lifelog platform.

\section{Activity tracking health care}
RSAT, the system to be developed is a tool intended to help physiotherapists and other clinicians to monitor a patient's physical activity. Its focus is to monitor over-weight patients with chronic kidney disease but it is assumed that it also could be used to monitor other groups of patients that for some reason needs to get more physical active.

The term AT health care will be used throughout this thesis to describe this kinds of treatments.

\section{Purpose and question formulation}
\label{sec:questions}

The purpose of this thesis is to build a system that can be used in the AT health care to help patients with chronic kidney disease to lose weight. The Sony Lifelog platform will be used as the foundation for the system. Therefore, an evaluation regarding how well suited the Sony Lifelog platform actually is for being used in the RSAT will be performed.

There exists many different activity tracking devices on the market. A goal with this thesis is therefore to investigate what different properties those have and to compile a summary about what properties that an activity tracking device ideally should have in order to be used in the AT health care.

@@

\begin{itemize}
    \item \emph{How well suited is the Sony Lifelog platform to be used in the AT heath care from a technical standpoint?}

    \item \emph{How well suited is RSAT, the system to be developed, to be used in the AT health care?}
    \item \emph{What properties should an activity tracking device ideally have in order to be used in the AT health care?}
\end{itemize}

\section{Method}% TODO: Den här kommer behöva moddas, efter att resten är klart. 
Before the development of RSAT could start, an interview with a physiotherapist from the RSAT-group was held. The wishes from the physiotherapist was used as input when the requirements was created. Thereafter a proof-of-concept system was developed to verify that it was possible to use the Lifelog API.

Then iterative prototyping was used to develop RSAT. For each iteration, the current requirements was identified and implemented. Each iteration resulted in a demo that the physiotherapist evaluated. Feedback from the physiotherapist was used as input to the next iteration. Sometimes the feedback resulted in new requirements on the system and if those were feasible, they were added to a development iteration.

@@

\section{Disposition} %TODO SKRIVA OM. Förklara att bakgrundskapitlet om njursvikt kan hoppas över för den som bara ät tekniskt intresserad. Också förkalra att det ger en förklaring till varför vårt syfte är som det är. En idé vore också att ta bort bakgrundskapitlet. 
The thesis is organized into three different parts; an overview part, a development part and an evaluation part.

The overview part consists of chapters 2 and 3. It gives a medical background, describing why RSAT is needed and a technical background that describes the Lifelog platform.

%It also contains an introduction to PalCom, the framework that RSAT is built on. The last chapter in this part describes related work. Related work includes information about different wearable technologies, platforms for aggregation of activity data and other systems that are similar to RSAT.

The development part consists of chapters 4 and 5. This part contains information about the requirements on the system, what decisions that were made, why the decisions were made and how RSAT was implemented.

The evaluation part consists of chapters 6, 7 and 8. In chapter 6 RSAT is compared to system similar to it self. In the following chapters RSAT and Lifelog are evaluated and our conclusions are drawn.

\section{Contribution statement}

Most of the work in the thesis was done collectively by the two parties but in some cases one party was contributing more and the other. Jens did most of the Lifelog service. Alfred designed the databases and most of the database services.

%TODO! Skriva vem som gjort vad. 
%Länkar till exjobb som vi kan titta på: sam.cs.lth.se/ExjobGetFile?id=652, http://sam.cs.lth.se/ExjobGetFile?id=668

@@

In this chapter an introduction to chronic kidney disease is given, beginning with a short introduction to chronic conditions in general. Followed by a discussion about chronic kidney disease in particular.

@@

The costs in the health care are ever increasing. A huge part of the health care and the economic costs associated with it addresses patients with chronic conditions. More than 60\% of all deaths in the world is due to this types of complications. In Sweden 80\% of all health care costs go to treating these conditions \cite{SmartCare}. A study shows that three types of lifestyle behaviors  leads to four chronic diseases. These diseases accounts for about 50\% of all the deaths worldwide, see figure \ref{fig:chronic-fig} \cite{callaway2015quantified}.

@@

Physical inactivity, as seen in figure \ref{fig:chronic-fig}, is one factor that could lead to a chronic condition. An increased amount of physical activity can therefore lower the risks of these conditions.

\section{Chronic kidney disease}
\label{sec:kiney}
\begin{comment}
NKDEPBASICS: http://nkdep.nih.gov/learn/kidney-disease-basics.shtml
KIDNEYFOUNDATION: https://www.kidney.org/kidneydisease/aboutckd
NKDEPPROG: http://nkdep.nih.gov/identify-manage/manage-patients/slow-progression.shtml
DIALYSIS: http://www.vardhandboken.se/Texter/Dialys-peritonealdialys/Oversikt/
\end{comment}
Kidneys have the responsibility of filtering waste products from the blood in the body \cite{NKDEPBASICS}. Chronic kidney disease (CKD) is a gradual loss of kidney function over time \cite{KIDNEYFOUNDATION}. There are no specific symptoms of the disease but illness, tiredness and a reduced appetite are all common symptoms. As the kidney function is worsening, waste products build to high levels in the blood and may cause complications such as high blood pressure,  muscle and bone mass loss. The risk of getting other diseases such as cardiovascular disease, pericarditis and anemia is also growing.  When the waste products in a patient’s body becomes so high that the patient starts to get sick from it, the patient requires dialysis \cite{KIDNEYFOUNDATION}.%todo: kolla över mening

People with severe CKD are in need of a kidney transplant. Until a transplant can be performed, dialysis is required. Without dialysis, the patients would have a poor life expectancy \cite{KIDNEYFOUNDATION}. Dialysis is a cumbersome treatment process for a patient as it takes up a lot of time and adds a lot of constraints to the patient’s life. Dialysis is also expensive for the society. A rough estimate of the annual cost of treating one patient with dialysis in Sweden is 250 000 – 500 000 SEK \cite{sr}.

Peritoneal dialysis and hemodialysis are dialysis that can be performed in the home of a CKD-patient. The principle behind peritoneal dialysis is that a sterile solution containing glucose is run into the patient’s abdomen \cite{DIALYSIS}. The peritoneum inside the patient acts as a membrane across which waste products from the blood is filtered. A downside with the peritoneal dialysis is that the calorie intake increases, since the sterile solution contains glucose. Hemodialysis is a kind of blood dialysis, where blood is filtered outside the body.

Despite that CKD is considered to be an irreversible condition, the progress of the disease can be slowed down. The National Kidney Disease Education Program (NKDEP) lists three strategies that can be used to slow down the progress \cite{NKDEPPROG}. One of those strategies is lifestyle interventions and addresses different health-promoting behaviour. One of the health-promoting behaviours they discuss is physical activity. NKDEP states that people suffering from CKD have the same recommendations for physical activity as the general population.

Eventually, a patient with CKD must undergo a kidney transplant. The patient must be in good physical shape, as the risk of infections and other diseases otherwise are too high. It is also essential that the patient does not have too much belly fat. Not only does the fat induce a greater risk for the patient's life, it also makes it harder for surgeons to practically perform the transplant. It is in the phase RSAT wants to help clinician to monitor patients and help them loose weight.

%Vad kostar det?!
%http://www.kidney.ca/document.doc?id=4083
%http://health.costhelper.com/kidney-cancer.html

\chapter[Wearable Technology]{Wearable Technology}

This chapter begins with a description and discussion about wearable technology, describing SWR10(the smart band used to collect movement data from patients in this thesis) and the context it operates in. Following by a discussion about platforms for saving data collected from wearable technology.

\section{Smart wristbands in general}

Wearable health and fitness devices are becoming more and more popular. Smart wristbands are the most popular devices  in this category, though it is predicted that the smart watches will become favored over the wristbands in 2015 \cite{gartner}. The difference between a smart wristband and a smartwatch is a bit of a grey area. A smart wristband and a smartwatch have more or less the same functionality. Smart watches in general tend to have more functionality than a wristband, but this is not always the case.

%There exists a wide range of wearable health and fitness devices. A new technology in this area is "smart clothing", where different garments can track data such as movements and heart rate\cite{callaway2015quantified}. Progress has also been made in the fields of biosensors. Scientists at the University of San Diego have developed a tattoo-like sensor, which can measure the lactate levels in the blood by converting sweat into electricity\cite{tattoo-device}.

%\subsection{Smart wristbands in general}
%\label{sec:Wearable}
The focus in this section is on wristbands but the descriptions also holds for smart watches. For this thesis the SWR10 is the chosen activity tracker. This is done because SONY is a stakeholder in the project and has chosen that smart band for the project.

\subsection{Accelerometer}
Movement tracking and sleep tracking are two common features in the smart wristbands that exist on the market. According to Quantified Wellness, an article published by RGA Reinsurance Company \cite{callaway2015quantified}, most of the data are collected through an accelerometer in the device. Many of the smart wristbands also takes advantages of the GPS in the smartphone they are connected to and thereby increases the accuracy of the measurements.

The Sony SmartBand SWR10 uses an accelerometer for recognizing different movements a user makes \cite{SWR10}.

%The SWR10 was chosen beacuse SONY is a stakeholder in the project and has decided the smart band fore the project.

For use in the AT heath care it is essential that the patients and the clinician can trust the data gathered by the wristband to be able to draw conclusions from it. 
Data collected in this type of fitness devices has shown to differ a lot. For example an equally long walk could differ between activitys tracker from 8.9 km to 14.5 km \cite{wearable-technology-cannot-be-trusted}.
Studies about the accuracy of the SWR10 has not been found in the research process of this thesis and due to other priorities a field study of this was not conducted in the thesis.

\subsection{Other sensors}
Other features that exist in wristbands on the market are the ability to track a user's heart rate, body temperature, blood oxygen level, respiration rate and galvanic skin response. The ways these sensors are used in different wristbands differ. In some wristbands, the user must activate a sensor in order to make a measurement and in some wristbands data are collected without any interaction with the user at all.

The possibility to see the patients heart rate is something the physiotherapist in the RSAT group expressed an interest for. If the heart rate was known, it would be easier to know how active each patient is during a day. The fact that the SWR10 does not have any heart rate sensor is therefore one of its major disadvantages.

%The most common type of heart rate sensor is optical and uses LED-lights in combination with an optical sensor to detect the amounts of blood flowing through the wrist in order to calculate a user’s pulse. The accuracy of different optical heart rate sensors varies a lot. In a test performed by CNET, four fitness trackers with an optical heart rate sensor were used to measure the heart rate. The results were then compared to the pulse measured by an EKG machine. The variance in the results were big.  At a pulse rate of 80 – 90 bpm, all of the devices was able to measure the heartbeat and was at the most 10\% off from the real heart rate. At a heart rate of 160 – 170 bpm, one of the trackers could not measure any data at all and two of the trackers were more than 50\% off\cite{CNET}. One of the devices used in the test, a Samsung Galaxy S5 smartphone, proved to be very accurate. It was at most 3\% off compared to the EKG-result. A chest strap was also used to measure the pulse. The results from the chest strap was very close to the EKG when the heart rate was between 160-170 and about 10\% off when it measured the heart rate at 80 – 90 bpm. The chest strap measures the pulse by measuring electrical pulse in a similar way to the EKG-machine and it is therefore expected that the results from the chest strap should be more accurate.

%Another type of heart rate monitor used in some wristbands is an impedance sensor. The impedance sensor measures the resistance of body tissue and uses that data to calculate the pulse. Jawbone Up3 is using this technique\cite{up3}. These devices have recently hit the market and any comments about their actual accuracy can therefore not be given. The pro with the impedance sensor is that it can be used to measure more data such as respiration rate and galvanic skin response. Important to mention is that Up3 and Up4 will mainly focus on resting heart rate. This is something that most of the manufacturers of fitness trackers that are using optical sensors also state that they do.

\subsection{Display}
\label{sec:display}
Some wristbands on the market are equipped with a display. The display can be used to deliver different kinds of intelligence feedback to the user and also to deliver different kinds of notifications such as if the user has been inactive for too long or reached an activity goal.  Feedback and notifications are also delivered to the user through different kinds of vibrations and sounds in the wristbands. A display makes the smart bands more independent of the smartphone. A smartband that has a display only needs the smartphone for synchronizing data to the cloud. Real time update is not considered to be an important feature in RSAT meaning that synchronizing the bands once a day would be acceptable. A smartband with a display would therefore eliminate the need for patients to always carry the smartphone with them wherever they go as they could get sufficient feedback about their physical activities from the smart band.

\subsection{Smartphone dependency}
Most or all of the smart wristbands needs to be connected to a smartphone. The synchronization to the smartphone is performed using Bluetooth. Many of the wristband providers also synchronize the data to their own web servers, where the data sometimes is accessible through an API. Some of these API:s are free to use and for some API:s a licence is needed.

Most of the smart wristbands on the market are strongly tied to the manufacturer’s own ecosystem, meaning that a user needs to use the manufacturers own applications to view the data collected by the wristband. This is also the case with SWR10. This application is called Lifelog.

The Sony Lifelog application is an Android application which is built around the idea of logging a users lifestyle and daily activities \cite{LifeLogDescr}. The application collects various data about its users through different sensors in a smartphone and SmartWear-devices (Sony SmartWatch or a Sony Smartband) and displays it to the user. Examples of data it logs are information about what applications in the smartphone that have been used, what locations that have been visited and information about different physical activities performed during the day. This application collects the information from the SWR10.

An interesting company, which has chosen to not make there own application, is Angel. Angel has, according to themselves \cite{angelfaq}, developed the first smart wristband for health and fitness. It is based on an open platform, which includes connectivity, protocols, and SDK.

% Jens kolla jag tog bort detta
%\subsection{Usage}

% The awareness and usage of wearables are on the rise. However, it has been showed that a third of all users who obtained a fitness device more than 12 months ago are no longer using it \cite{callaway2015quantified}.

%This chapter describes the techniques used to develop RSAT. First the Sony Lifelog and techniques used by Lifelog are described. The second part describes PalCom which is the platform used by the itACiH and RSAT.

%\subsection{Sony SmartBand}

%The Sony SmartBand SWR10 is the chosen wristband for this project. This is because SONY is a stakeholder in the project and has decided the smart band fore the project. The Sony SWR10 has a movement sensor that recognizes different movements a user makes\cite{SWR10}. This data is analyzed by Sony and categorized into different activates, which are uploaded to a central repository that stores the data. The SmartBand must be connected to an Android device with Android version 4.4 or higher to be able to collect data.

% It is essential that the patients and the physiotherapist can trust the data that is gathered by the wristband to be able to draw conclusions from it. Studies about the accuracy of the SWR10 have not been found in the research process of this thesis and due to other priorities a study of this was not conducted in the thesis.

%\subsubsection{Sony Lifelog}

%The Sony Lifelog application is an Android application which is built around the idea of logging a users lifestyle and daily activities\cite{LifeLogDescr}. The application collects various data about its users through different sensors in a smartphone and SmartWear-devices (Sony SmartWatch or a Sony Smartband) and displays it to the user. Examples of data it logs are information about what applications in the smartphone that have been used, what locations that have been visited and information about different physical activities performed during the day. This application collects the information from the SWR10.

%\subsubsection{Accuracy}

%To get accurate data from activity trackers there are essentially two important things: (1) - to get good sensor data and (2) - good algorithms to analyze the sensor data. Studies about the quality of the sensor data and the algorithms for SWR10 have not been found in the research process of this thesis.

%There exist other wristbands on the market that have more sensors. More sensors could imply that more accurate data could be collected. This is however, something that highly depends on the accuracy of the sensor data and the algorithms used. It is essential that the patients and the physiotherapist can trust the data that is gathered by the wristband to be able to draw conclusions from it. More sensors should therefore not be included if there is a risk that the collected data is inaccurate. More sensors would not only mean that it was possible to collect more data, it would also mean that calculations about things such as physical activity could be improved since more data could be used as input to the calculations. %todo utveckla med exempel kanske?

%A heart rate is something the physiotherapist has expressed an interest of being able to see. If the heart rate was known, it would be easier to know how active each patient is during a day. The fact that the SWR10 does not have any heart rate sensor is therefore one of its major disadvantages. On the other hand, it is good that Sony has chosen to not integrate a sensor that cannot deliver useful data. As stated before, optical heart rate sensors tend to perform badly at higher heart rates. The usefulness of optical heart rate monitors when exercising is therefore somewhat debatable. But the data from a heart rate monitor could still be used as input to find out whether somebody performs any physical activity or is inactive. This would be valuable information for the physiotherapist.

%Some of the smart bands have a display and this makes the smart bands more independent of the smartphone. A smart band that has a display only needs the smartphone to synchronize data to the cloud. Real time update is not an important feature in RSAT meaning that syncronizing the bands once a day would be acceptable. Therefore a display could eliminate the need for patients to always carry both a smartband and a smartphone.

\section{Different Activity Tracking Platforms}
This section first describes and evaluates the Sony lifelog platform. Sony lifelog is thereafter compared to other platforms from a developers perspective.

\subsection{Sony Lifelog API} 
All the data collected in the Sony Lifelog application are automatically uploaded to a central repository in the Sony cloud that stores the data. This data is globally accessible through a REST-based web-service API \cite{LifeLogDescr}.

@@

Third parties can access the API to retrieve collected data about a user. Every request made to the API must be authorized and https must be the chosen communication protocol. The lifelog Web API uses OAuth 2.0 as its authorization mechanism \cite{LifeLogDescr}. To connect to the API, a user must login and give explicit permission for the application or the service to access to the data.

@@

The Sony Lifelog API claims to be a REST API. REST is an architectural style for designing API:s and are heavily used in modern web services. The concept is to represent resources as the analogy for the data endpoints and to have a small set of actions interacting with these resources. The REST architecture is based on the following principles \cite{pautasso2014restful}:

@@

Sony states that the Lifelog API is REST-based. In order to be a REST API the API must fulfill the architectural principles that presented above. We have evaluated the API and concluded that it actually is a REST API. Below you find comments on why the API fulfils each principle:

\begin{itemize}
    \item \emph{Addressability} - Each request to the Sony Lifelog contains a time interval and a list of data types telling what you want your response to contain. The queries to the API are idempotent. Two queries containing the same request will always be responded with equal results. The addressability principle therefore holds, as you can address a resource using a unique URL. 
    \item \emph{Uniform interface} - The Sony Lifelog is a read-only API and only supports GET requests. It therefore has a uniform interface.
    \item \emph{Stateless Interaction} -  The communication with the API is stateless. The only exception to this is the authentication part of the communication.
    %that is using OAuth2. When an access token is issued, the API must verify that the client’s credentials are correct and the authentication is therefore not stateless.
    However, the authentication can be seen as something that happens before the connection to the API, which would make the API-communication completely stateless.
    \item \emph{Self-describing Messages} - The strings used to query the API are self-describing and uses distinct parameters. The responses have a JSON-syntax, are self-describing and have distinct parameters. 
    \item \emph{Hypermedia} - The hypermedia principle is used by the API when the requested data is too large to fit in a single response. The API uses links that it sends with the first part of the response. These links contains the URL to the rest of the queried data.
\end{itemize}

@@

The Sony Lifelog API uses OAuth2.0 as its authorization mechanism. A user must log in and give explicit permission for access to their data. When that is done, an application can get OAuth2.0 authorization for API calls that are made on that user’s behalf.

OAuth 2.0 is an authorization protocol developed by the Internet Engineering Task Force (IETF). It describes an open standard for authorization and enables third party applications to obtain limited access to a web-service on the behalf of a resource owner \cite{oauth2spec}. To describe the protocol, several key actors need to be described, these actors are the client, the resource owner, the authorization server and the resource server. IETF gives the following definitions to the roles:

@@

% Kolla källa nedan, kanske nämna "oauth2spec" igen

The Authorization Grant is a credential representing the resource owner's authorization. The IETF's OAuth 2.0 specification defines four different grant types; authorization code, implicit, resource owner password credentials and client credentials. The Sony Lifelog API uses the authorization code as its grant type and the other types will therefore not be further discussed \cite{LifeLogAuth}. When the authorization code grant type is used, the resource owner is redirected to an authentication server when the client requests authentication. When the resource owner has granted the request, the resource owner is redirected back to the client together with the authorization code, which is appended in the URL. The authentication code is short lived and single-use.

@@

%\section{Evaluation of RSAT}
